Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers

### 2025-01-24 endofquarterpaper.qmd
edits

~One of the most~ A distinctive quality~ies~ of the Upper Paleolithic is the large rise in ~available~ artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans <!--BM: scholarly citations needed here-->. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe <!--BM: scholarly citations needed here-->. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

@@

Multivariate analyses of Gravettian beads from burial and occupation sites in Europe suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis ~are~ were <!--BM: everything in this background section needs to be in past tense, they have already done their work and it is complete --> used to identify the groups. The geographically discrete clusters identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions. <!--BM: this is a 3 sentence paragraph, that's a sign that something is wrong. Either add more detail, or combine with another para. I think adding more detail is ok here -->

@@

Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done. The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans. As said previously, previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.

@@

@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons. The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted. Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups. This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move. Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas. Again, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.

>>>>

~One of the most~ A distinctive quality~ies~ of the Upper Paleolithic is the large rise in ~available~ artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans <!--BM: scholarly citations needed here-->. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe <!--BM: scholarly citations needed here-->. Considered to be the starting point of the Upper Paleolithic, the Aurignacian ~technocomplex~ period (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

@@

Multivariate analyses of Gravettian beads from burial and occupation sites in Europe suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis ~are~ were <!--BM: everything in this background section needs to be in past tense, they have already done their work and it is complete --> used to identify the groups. The geographically discrete clusters identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions. <!--BM: this is a 3 sentence paragraph, that's a sign that something is wrong. Either add more detail, or combine with another para. I think adding more detail is ok here. What was their data, what were their methods, what were their results, in detail. ratio of words to information, mention 'isolation-by-distance' -->

@@

Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done. The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans. ~As said previously~ <!--BM: no, we write so logically  that we don't need this -->, previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.

@@

~@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons. The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted. Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups. This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move. Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas. Again, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.~ <!--BM: better in the discuss as context and interpretation -->
